Simulating Ultra High-Dimensional Multivariate Data Using the bigsimr R Package

### 2020-09-21 010-background.Rmd
clarify vision and organize

We present an general-purpose, scalable multivariate simulation algorithm. The crux of the method is construction of a Gaussian copula. This idea is well known [refs]. A copula is a distribution function on $[0,1]^d$ describing a random vector with standard uniform marginals. Moreover, for any random vector ${\bf X}=(X_1, \ldots, X_d)$ with cumulative distribution function (CDF) $F$ and marginal CDFs $F_i$ there is a copula function 
$C(u_1, \ldots, u_d)$ so that

>>>>

We present an general-purpose, scalable multivariate simulation algorithm. The crux of the method is construction of a Gaussian copula. This idea is well known [refs]. We chose a Gaussian copula among the many available copula functions primarily for two reasons: 1) the computationally convient closed expression to convert among the most common dependency measures when the margins are Gaussian and 2) the availability of high-performance multivariate normal simulators in R ('mvnfast' ref).

A copula is a distribution function on $[0,1]^d$ describing a random vector with standard uniform marginals. Moreover, for any random vector ${\bf X}=(X_1, \ldots, X_d)$ with cumulative distribution function (CDF) $F$ and marginal CDFs $F_i$ there is a copula function 
$C(u_1, \ldots, u_d)$ so that

### 2020-09-21 030-bigsimr-package.Rmd
clarify vision and organize

```{r setup3, eval = TRUE}
## setwd('~/Research/article_bigsimr/'); bookdown::preview_chapter('030-bigsimr-package.Rmd')
## bookdown::render_book('index.html')
library(bigsimr)
library(reticulate)
reticulate::use_condaenv("bigsimr-cpu")
```

@@

```{r setMargins}
margins = alist(
  qnorm(mean = 3.14, sd = 0.1),
  qbeta(shape1 = 1, shape2 = 4),
  qnbinom(size = 10, prob = 0.75)
)
```

@@

```{r setCor}
rho <- matrix(0.5, nrow = 3, ncol = 3)
diag(rho) <- 1.0
rho
```

@@

```{r rvec10}
x <- bigsimr::rvec(10, rho = rho, margins = margins, type = "spearman")
```

@@

```{r printX}
x
```

@@

```{r rvec1e6, fig.width=7}
x <- rvec(1e6, rho = rho, margins = margins, type = "spearman")

@@

```{r fastCor}
bigsimr::fastCor(x, method = 'spearman')
```

@@

```{r computeBounds, echo = TRUE, eval=FALSE}
## Possibly bug? or configuration
## Error: segfault from C stack overflow on my Okapi bigsimr env 
rho_bounds <- bigsimr::computeCorBounds(margins = margins, type = 'spearman', cores = 1, reps = 1e5)
rho_bounds
```

@@

```{r rvecRslurm, echo=TRUE, eval=FALSE}
library(rslurm)
## a single call of rvec
## had to edit rslurm/templates/submit_single_sh.txt
## source ~/.bashrc
## conda activate bigsimr
## conda env config vars list
sjob <- slurm_call(rvec, jobname = 'rvec',
                   list(n=1e6,
                        rho = rho,
                        margins = margins,
                        type = "spearman"),
                   submit = TRUE)
```

```{r rvecRslurmRes, echo=TRUE, eval=FALSE}
res <- readRDS(file.path('_rslurm_rvec', list.files('_rslurm_rvec', 'results')))
head(res, 3)
## cleanup_files(sjob) ## careful this deletes output also
```

@@

```{r rvecRslurmMap, echo=TRUE, eval=FALSE}
library(rslurm)
## a single call of rvec
## had to edit rslurm/templates/submit_sh.txt
## source ~/.bashrc
## conda activate bigsimr
## conda env config vars list
set.seed(06202020)
simReps <- 100
rhoList <- replicate( n = simReps, bigsimr::rcor(d = length(margins), constant_rho = TRUE), simplify=FALSE )
sjob <- slurm_map(x = rhoList,
                  f = rvec,
                  jobname = 'rvecMap',
                  n=1e6,
                  margins = margins,
                  type = "spearman",
                  nodes = 4,
                  cpus_per_node = 4,
                  submit = TRUE)

@@

```{r rvecRslurmMapRes, echo=TRUE, eval=FALSE}
list.files('_rslurm_rvecMap', 'results')
res <- rslurm::get_slurm_out(sjob, outtype = 'raw')
rhoHats <- lapply( res, bigsimr::fastCor, method = 'spearman' )

@@

```{r rvecRslurmMapFig, echo = TRUE, eval=TRUE, message=FALSE, warning=FALSE, out.width= '75%', fig.align='center', fig.cap='The continous marginal pairs exact reproduce the specified correlation whereas the discrete pairs (green nad blue) show a downward bias. Future work with employ a discrete adjusted procedures using *rescaling* on the traditional correlation coefficient.'}
library(tidyverse)

>>>>

```{r setup3, eval = FALSE}
## setwd('~/Research/article_bigsimr/'); bookdown::preview_chapter('030-bigsimr-package.Rmd')
## bookdown::render_book('index.html')
library(bigsimr)
library(reticulate)
reticulate::use_condaenv("bigsimr-cpu")
```

@@

```{r setMargins, eval = FALSE}
margins = alist(
  qnorm(mean = 3.14, sd = 0.1),
  qbeta(shape1 = 1, shape2 = 4),
  qnbinom(size = 10, prob = 0.75)
)
```

@@

```{r setCor, eval = FALSE}
rho <- matrix(0.5, nrow = 3, ncol = 3)
diag(rho) <- 1.0
rho
```

@@

```{r rvec10, eval = FALSE}
x <- bigsimr::rvec(10, rho = rho, margins = margins, type = "spearman")
```

@@

```{r printX, eval = FALSE}
x
```

@@

```{r rvec1e6, fig.width=7, eval = FALSE}
x <- rvec(1e6, rho = rho, margins = margins, type = "spearman")

@@

```{r fastCor, eval = FALSE}
## help( package = 'bigsimr')
bigsimr::cor_fast(x, method = 'spearman')
```

@@

```{r computeBounds, echo = FALSE, eval=FALSE}
## Possibly bug? or configuration
## Error: segfault from C stack overflow on my Okapi bigsimr env 
rho_bounds <- bigsimr::computeCorBounds(margins = margins, type = 'spearman', cores = 1, reps = 1e5)
rho_bounds
```

@@

```{r rvecRslurm, echo=FALSE, eval=FALSE}
library(rslurm)
## a single call of rvec
## had to edit rslurm/templates/submit_single_sh.txt
## source ~/.bashrc
## conda activate bigsimr
## conda env config vars list
sjob <- slurm_call(rvec, jobname = 'rvec',
                   list(n=1e6,
                        rho = rho,
                        margins = margins,
                        type = "spearman"),
                   submit = TRUE)
```

```{r rvecRslurmRes, echo=FALSE, eval=FALSE}
res <- readRDS(file.path('_rslurm_rvec', list.files('_rslurm_rvec', 'results')))
head(res, 3)
## cleanup_files(sjob) ## careful this deletes output also
```

@@

```{r rvecRslurmMap, echo=FALSE, eval=FALSE}
library(rslurm)
## a single call of rvec
## had to edit rslurm/templates/submit_sh.txt
## source ~/.bashrc
## conda activate bigsimr
## conda env config vars list
set.seed(06202020)
simReps <- 100
rhoList <- replicate( n = simReps, bigsimr::rcor(d = length(margins), constant_rho = TRUE), simplify=FALSE )
sjob <- slurm_map(x = rhoList,
                  f = rvec,
                  jobname = 'rvecMap',
                  n=1e6,
                  margins = margins,
                  type = "spearman",
                  nodes = 4,
                  cpus_per_node = 4,
                  submit = TRUE)

@@

```{r rvecRslurmMapRes, echo=FALSE, eval=FALSE}
list.files('_rslurm_rvecMap', 'results')
res <- rslurm::get_slurm_out(sjob, outtype = 'raw')
rhoHats <- lapply( res, bigsimr::fastCor, method = 'spearman' )

@@

```{r rvecRslurmMapFig, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, out.width= '75%', fig.align='center', fig.cap='The continous marginal pairs exact reproduce the specified correlation whereas the discrete pairs (green nad blue) show a downward bias. Future work with employ a discrete adjusted procedures using *rescaling* on the traditional correlation coefficient.'}
library(tidyverse)

### 2020-09-21 050-some-applications.Rmd
clarify vision and organize

```{r LoadLib050, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
library(ggplot2)
library(tidyverse)
library(knitr)
library(dplyr)
## devtools::install_github( "adknudson/bigsimr", ref = 'develop')
library(bigsimr)
reticulate::use_condaenv("py37")
CORES <- parallel::detectCores() - 1
set.seed(06082020)
## devtools::install_github("ggobi/GGally")
## devtools::install_github( "adknudson/bigsimr")
## A conda environment is also activitied (see README)
library(bigsimr)
library(GGally)
```

@@

```{r processBRCA, echo=TRUE, eval=TRUE, cache=TRUE}
## full workflow simulating RNA-seq data
allDat <- readRDS( file = "~/Downloads/complete_processed_tcga2stat_RNASeq2_with_clinical.rds" )
lastClinical <- which( names(allDat) == 'tumorsize' )
brca <- allDat[ allDat$disease == "BRCA", (lastClinical+1):ncol(allDat) ]
## remove naively the RSEM adjustment
brca <- round(brca, 0)
ncol(brca) ## num of genes 20501
## compute the median expression! avoid the 0s
brcaMedian <- apply(brca, 2, median)
## retain top (1-probs)*100% highest expressing genes for illustration
myProb <- 0.99
cutPoint <- quantile( x = brcaMedian, probs = myProb )
## genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
brca <- brca[ , genesToKeep ]
## ncol(brca) / 20501
(d <- length(genesToKeep))
```

@@

```{r figNBratio, echo = TRUE, eval=TRUE, message=FALSE, warning=FALSE, out.width= '75%', fig.align='center', fig.cap='The genes display extreme overdispersion. Using group-specific sample means and variances for each of the d=`r d` genes, we display the distributions oflog of variance to mean ratio.'}

@@

```{r estRhoBRCA, echo=TRUE, eval=TRUE}
## full workflow simulating RNA-seq data

@@

```{r estMargins, echo = TRUE, eval = TRUE, cache=TRUE}
## 2. Estimate NegBin parameters using Method of Moments
estimateNegBinMoM <- function(tmpGene) {
    tmpMean <- mean(tmpGene)
    tmpVar <- var(tmpGene)
    ## relate to nbinom parameters
    ## See ?rbinom for details.
    p <- tmpMean / tmpVar
    n <- ( tmpMean * p) / ( 1  - p )  
    ## format for bigsimr margins
    get( 'qnbinom')
    return( str2lang( paste0( "qnbinom(size = ", n, ", prob = ", p, ")" ) ) )
}
brcaMargins <- apply( unname(as.matrix(brca)), 2, estimateNegBinMoM )
## brcaMargins <- apply( unname(as.matrix(brca))[ , 1:3], 2, estimateNegBinMoM )
## describe the margins
## check for too small prob
## head( sort( unlist( lapply( brcaMargins, function(x) {x$prob} ) ) ) )
## head( sort( unlist( lapply( brcaMargins, function(x) {x$size} ) ) ) )
```

```{r simBRCA, echo=TRUE, eval = TRUE, cache=TRUE}
## 3. Generate the simulated samples
N <- nrow(brca)
system.time( simBRCA <- rvec(n = N, rho = rho, margins = brcaMargins, cores = CORES, type = corType) )
simBRCA[1:2, 1:2 ]

@@

```{r evalBRCA, echo=TRUE, eval = FALSE, cache=TRUE}

@@

```{r densityEvaluation, echo=TRUE, eval=FALSE}
## full workflow simulating RNA-seq data

>>>>

```{r LoadLib050, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, results = "hide"}
library(ggplot2)
library(tidyverse)
library(knitr)
library(dplyr)
## devtools::install_github( "adknudson/bigsimr", ref = 'develop')
library(bigsimr)
reticulate::use_condaenv("bigsimr")
CORES <- parallel::detectCores() - 1
set.seed(06082020)
## devtools::install_github("ggobi/GGally")
## devtools::install_github( "adknudson/bigsimr")
## A conda environment is also activitied (see README)
library(bigsimr)
library(GGally)
```

@@

```{r processBRCA, echo=FALSE, eval=FALSE, cache=TRUE}
## full workflow simulating RNA-seq data
allDat <- readRDS( file = "~/Downloads/complete_processed_tcga2stat_RNASeq2_with_clinical.rds" )
lastClinical <- which( names(allDat) == 'tumorsize' )
brca <- allDat[ allDat$disease == "BRCA", (lastClinical+1):ncol(allDat) ]
## remove naively the RSEM adjustment
brca <- round(brca, 0)
ncol(brca) ## num of genes 20501
## compute the median expression! avoid the 0s
brcaMedian <- apply(brca, 2, median)
## retain top (1-probs)*100% highest expressing genes for illustration
myProb <- 0.99
cutPoint <- quantile( x = brcaMedian, probs = myProb )
## genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
brca <- brca[ , genesToKeep ]
## ncol(brca) / 20501
(d <- length(genesToKeep))
```

@@

```{r figNBratio, echo=FALSE, eval=FALSE, message=FALSE, warning=FALSE, out.width= '75%', fig.align='center', fig.cap='The genes display extreme overdispersion. Using group-specific sample means and variances for each of the d=`r d` genes, we display the distributions oflog of variance to mean ratio.'}

@@

```{r estRhoBRCA, echo=FALSE, eval=FALSE}
## full workflow simulating RNA-seq data

@@

```{r estMargins, echo=FALSE, eval=FALSE, cache=TRUE}
## 2. Estimate NegBin parameters using Method of Moments
estimateNegBinMoM <- function(tmpGene) {
    tmpMean <- mean(tmpGene)
    tmpVar <- var(tmpGene)
    ## relate to nbinom parameters
    ## See ?rbinom for details.
    p <- tmpMean / tmpVar
    n <- ( tmpMean * p) / ( 1  - p )  
    ## format for bigsimr margins
    get( 'qnbinom')
    return( str2lang( paste0( "qnbinom(size = ", n, ", prob = ", p, ")" ) ) )
}
brcaMargins <- apply( unname(as.matrix(brca)), 2, estimateNegBinMoM )
## brcaMargins <- apply( unname(as.matrix(brca))[ , 1:3], 2, estimateNegBinMoM )
## describe the margins
## check for too small prob
## head( sort( unlist( lapply( brcaMargins, function(x) {x$prob} ) ) ) )
## head( sort( unlist( lapply( brcaMargins, function(x) {x$size} ) ) ) )
```

```{r simBRCA, echo=FALSE, eval=FALSE, cache=TRUE}
## 3. Generate the simulated samples
N <- nrow(brca)
system.time( simBRCA <- rvec(n = N, rho = rho, margins = brcaMargins, cores = CORES, type = corType) )
simBRCA[1:2, 1:2 ]

@@

```{r evalBRCA, echo=FALSE, eval=FALSE, cache=TRUE}

@@

```{r densityEvaluation, echo=FALSE, eval=FALSE}
## full workflow simulating RNA-seq data

### 2020-09-21 index.Rmd
clarify vision and organize

--- 
title: "Simulating Ultra High-Dimensional Multivariate Data Using the bigsimr R Package"
author: [A.G. Schissler, R.D.L. Foote, A.D. Knudson, T.J. Kozubowski, A.K. Panorska]
date: "`r format(Sys.Date(), '%A, %B %d, %Y')`"
site: bookdown::bookdown_site
bibliography: [bigsimr.bib, packages.bib]
biblio-style: apalike
link-citations: yes
abstract: |
  In this era of Big Data, it is critical to realistically simulate data to conduct informative Monte Carlo studies. This is problematic when data are inherently multivariate while at the same time are (ultra-) high dimensional. This situation appears frequently in observational data found on online and in high-throughput biomedical experiments (e.g., RNA-sequencing). Due to the difficulty in simulating realistic correlated data points, researchers often resort to simulation designs that posit independence --- greatly diminishing the insight into the empirical operating characteristics of any proposed methodology.  Major challenges lie in the computational complexity involved in simulating these massive random vectors. We propose a fairly general, scalable procedure to simulate high-dimensional multivariate distributions with pre-specified marginal characteristics and dependency characteristics. As a motivating example, we use our methodology to study large-scale statistical inferential procedures applied to cancer-related RNA-sequencing data sets. The proposed algorithm is implemented as the `bigsimr` R package.
---

```{r indexSetup, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
library(ggplot2)
library(tidyverse)
library(knitr)
library(dplyr)
set.seed(06082020)
```

@@

Massive high-dimensional data sets are commonplace in many areas of scientific inquiry. As new methods are developed for these data, a fundamental challenge lies in designing and conducting simulation studies to assess operating characteristics of proposed methodology, such as false positive rates, statistical power, and robustness --- often in comparison to existing methods. Another application lies in the discovering the sampling distribution of test statistic under hypothesized null models in complex scenarios. Such Monte Carlo studies in the high-dimensional setting often become difficult to conduct with existing algorithms and tools, or require expertise in computing on clusters. This is particularly true when simulating massive multivariate, non-normal distributions.

>>>>

--- 
title: Simulating Ultra High-Dimensional Multivariate Data Using the bigsimr R Package
author:
  - name: Alice Anonymous
    email: alice@example.com
    affiliation: Some Institute of Technology
    footnote: 1
  - name: Bob Security
    email: bob@example.com
    affiliation: Another University
  - name: Cat Memes
    email: cat@example.com
    affiliation: Another University
    footnote: 2
  - name: Derek Zoolander
    email: derek@example.com
    affiliation: Some Institute of Technology
    footnote: 2
address:
  - code: Some Institute of Technology
    address: Department, Street, City, State, Zip
  - code: Another University
    address: Department, Street, City, State, Zip
footnote:
  - code: 1
    text: "Corresponding Author"
  - code: 2
    text: "Equal contribution"
date: "`r format(Sys.Date(), '%A, %B %d, %Y')`"
bibliography:
  - bigsimr.bib
  - packages.bib
site: bookdown::bookdown_site
journal: "Journal of Computational And Applied Mathematics (JCAM)"
#linenumbers: true
#numbersections: true
abstract: |
  In this era of Big Data, it is critical to realistically simulate data to conduct informative Monte Carlo studies. This is problematic when data are inherently multivariate while at the same time are (ultra-) high dimensional. This situation appears frequently in observational data found on online and in high-throughput biomedical experiments (e.g., RNA-sequencing). Due to the difficulty in simulating realistic correlated data points, researchers often resort to simulation designs that posit independence --- greatly diminishing the insight into the empirical operating characteristics of any proposed methodology.  Major challenges lie in the computational complexity involved in simulating these massive random vectors. We propose a fairly general, scalable procedure to simulate high-dimensional multivariate distributions with pre-specified marginal characteristics and dependency characteristics. As a motivating example, we use our methodology to study large-scale statistical inferential procedures applied to cancer-related RNA-sequencing data sets. The proposed algorithm is implemented as the `bigsimr` R package.
---

```{r indexSetup, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
library(tidyverse)
library(knitr)
library(rmarkdown)
library(bookdown)
## setwd('~/article_bigsimr')
## bookdown::render_book("index.Rmd", "bookdown::pdf_book")
## bookdown::serve_book()
set.seed(09212020)
```

@@

Massive high-dimensional data sets are now commonplace in many areas of scientific inquiry. As new methods are developed for these data, a fundamental challenge lies in designing and conducting simulation studies to assess operating characteristics of proposed methodology, such as false positive rates, statistical power, and robustness --- often in comparison to existing methods. Another application lies in the discovering the sampling distribution of test statistic under hypothesized null models in complex scenarios. Such Monte Carlo studies in the high-dimensional setting often become difficult to conduct with existing algorithms and tools, or require expertise in computing on clusters. This is particularly true when simulating massive multivariate, non-normal distributions.
